---
title: "Lecture 01"
subtitle: "Introduction to Data Science, R, and First Scripts"
format: revealjs
---

## Main Message {.main-message}

*Data Science is a multidisciplinary field that combines analytical thinking, programming, and storytelling to extract insights from data.*

*This course serves as a practical introduction to the field through the use of `R` and `RStudio` as main tools to think, code, and inspect data.*

# Course Introduction

## What is Data Science?

Data science is an interdisciplinary field that uses scientific methods, processes, algorithms, and systems to extract knowledge and insights from structured and unstructured data.

The data science lifecycle can be summarized:

- Starts from a concrete question.
- Data is collected and processed.
- Exploratory analysis finds patterns.
- Analytical models are applied to make explanations, predictions, or decisions.
- Ends with results that must be interpreted and communicated.

## What is a Data Scientist?

A *data scientist* is "a person who is better at statistics than any software engineer and better at software engineering than any statistician." - Josh Wills.

The core pillars and skills for Data Science are:

- **Analytics**: understanding the data.
- **Programming**: working with the data.
- **Communication**: showing the data.

## Course Structure and Dynamics (Part 1)

General:

- Course taught in English for all lectures and assignments.
- All materials and interactions through UCAB's **Modulo 7**.
- Course Whatsapp group for quick questions and announcements in Spanish (optional).
- Assignment deadlines at Sunday 11:59 PM.

Each week on Wednesday-Thursday:

- Publication of weekly course announcement at **Modulo 7**'s course page.
- Short overview of content, session links, and assignmnet reminders.

## Course Structure and Dynamics (Part 2)

Following Mondays 7:00-10:00 AM:

- Online lectures trhough **Zoom** with live coding and discussion.
- Single 10-15 min break around halfway mark.
- Attendance is recorded at random intervals.
- All lecture content and code will be on slides and scripts.
- Students to use `R` and `RStudio` to follow along and practice.
- Early sessions focus on core tools, data structures, and workflow.
- Later sessions expand toward analysis, visualization, and applied cases.

## Course Content (Part 1)

Current course edition is under construction, but the general structure will be as follows:

**Module 1:**

- Introduction to `R` and `RStudio`.
- Data Wrangling.

**Module 2:**

- Data Visualization.
- Exploratory Data Analysis.

## Course Content (Part 2)

**Module 3:**

- Introduction to Machine Learning.
- Linear Regression.
- Logistic Regression.
- Descicion Trees.
- Random Forests.
- Model Evaluation.
- Unsupervised Learning.

**Module 4:**

- Reproducible Research.
- RMarkdown and Quarto.

## Course Evaluations (Part 1)

Evaluations will combine ongoing work and formal assessment.

All grades are recorded on **M7** and are only rounded at the end of the semester. The final grade is a weighted average of all components, with the following breakdown:

**Individual:**

- Problem sets (22.5%): Weekly sets corresponding with last lecture. Set number of questions from random pool.
- Class participation (10%): Designated spaced for students to make comments, ask questions, and share insights during lectures.

*Note:* Depending on course progression weights are subject to change on a 5 percentage points margin.

## Course Evaluations (Part 2)

**Group** (3-4 students per group, chosen by students in first week):

- Group project deliverables (22.5%): Periodic checkpoints for group project, including proposal, data collection, analysis.
- Group project final report (32.5%): Final written report at the end of the semester, summarizing the project and findings.
- Group project presentation (12.5%): Final presentation of the project to the class, highlighting key insights and methods.

*Note:* Depending on course progression weights are subject to change on a 5 percentage points margin.

## Notes on AI policy

It is reasonable to expect that students will use AI tools to assist with their work, and that is not prohibited. However, the following guidelines apply:

- AI tools can be used for brainstorming, debugging, and learning, but not for generating final answers to assignments.
- All work submitted must be the student's own, and any use of AI tools must be disclosed in the assignment submission.
- The lecturer will provide mandatory prompts and guidelines for responsible AI use in assignments, and students are expected to follow them.
- The course will use AI detection tools to check for potential misuse, and any suspected cases will be reviewed according to the university's academic integrity policies.

# What are `R` and `RStudio`?

## What is `R`?

`R` was developed as an interactive environment for data analysis, not as a general-purpose programming language in the style of `C` or `Java`.

- One of the main workhorse languages for data science. The other being `Python`.
- It is especially strong for statistical analysis and visualization.
- It encourages reproducible work through scripts.
- It is widely used in academia, research, policy work, and quantitative analysis.
- High synergy with the type of work ususally done by Economists.
- Does't require previous coding experience to get started.
- It is free and open source.
- Very versatile through a vast array of libraries and packages.

## Advantages of `R`

- It is free and open source.
- It runs on all major operating systems.
- Scripts and data objects can be shared across platforms.
- It has strong statistical and visualization capabilities.
- It scripts allow easy reproducibility.
- Relatively robust skill demand for Researchers.
- It has a large, active user community.
- New methods often reach `R` quickly through packages.

## Disadvantages of `R`

- Easy to learn the basics, but harder to master.
- Command line interface can be intimidating for new users.
- Not a general-purpose programming language, so it can be less efficient for some tasks.
- Its syntax can feel idiosyncratic.
- It does not always follow the conventions of other languages.
- Finding the right package for a certain task can be difficult.
- Libraries dependencies could make tedious to mantain all tools up to date. At the mercy of the package developers.
- Poorly written code is hard to mantain and debug.
- It is less central than `Python` in some AI workflows.

## What is `RStudio`?

`RStudio` is an *integrated development environment (IDE)* for data science projects for `R`.

- It's an interface that makes it easier to write and run `R` code.
- It keeps scripts, console work, objects, files, plots, and help in one place.
- It requires `R` to work with `RStudio`, but not vice versa.
- Like `R`, it is free and open source.
- It will be the main interface we use in this course.
- Might not be as versatile as `VSCode` or `Jupyter`, but it is more focused on `R` and has a more intuitive interface for beginners.
- Lacks easy support for coding agents without CLI workarounds, but it is more than enough for the scope of this course.

## The Source Pane

::::: columns
::: {.column width="52%"}
This is where we write and save scripts.

- Scripts are text files with `R` code that we can run.
- It is the main place for real work.
- Scripts give us a record of the analysis.
- Non executable comments can be added with `#` to explain the code.
- Can be sectioned for ease of navigation.
- It lets us build a reproducible workflow instead of working only from memory.
- Lines of codes can be individually run from here.
:::

::: {.column width="48%"}
*Insert the fixed `RStudio` screenshot here and highlight the Source pane.*
:::
:::::

## The Console

::::: columns
::: {.column width="52%"}
This is where `R` evaluates commands immediately.

- It gives instant feedback.
- It is useful for quick checks and exploration.
- It is not a good substitute for a saved script.
- Not recommended for long-term work because it is not saved and can become disorganized.
:::

::: {.column width="48%"}
*Insert the fixed `RStudio` screenshot here and highlight the Console.*
:::
:::::

## The Environment and History

::::: columns
::: {.column width="52%"}
This is where we inspect what has happened in the current session.

- The Environment shows the objects currently stored in the workspace.
- The History shows previously executed commands.
- Provides reasonable control over elements that can be reused and modified.
- Keeps track of session status.
- Additional tabs here can show connections, git status, and more.
:::

::: {.column width="48%"}
*Insert the fixed `RStudio` screenshot here and highlight the Environment and History area.*
:::
:::::

## Files, Plots, Packages, and Help

::::: columns
::: {.column width="52%"}
This area supports the rest of the workflow.

- Files help us navigate the working directory.
- Plots display visual output.
- Packages show installed libraries.
- Help gives access to documentation.
:::

::: {.column width="48%"}
*Insert the fixed `RStudio` screenshot here and highlight this pane.*
:::
:::::

## Intervention Space {.intervention-slide .incremental}

Before touching code, answer these two questions.

- Why is it risky to work only in the console?
- What is the practical difference between `R` and `RStudio`?

# First Codes

## Essential Shortcuts

There are a few useful shortcuts when working with scripts:

- `Ctrl + Enter`: run the current line or selection.
- `Ctrl + Shift + Enter`: run the entire script.
- `Ctrl + Shift + C`: comment or uncomment code (`#`). 
- `Ctrl + Shift + R`: create a section.
- `Alt + O`: collapse sections.

## Basic operations (Part 1)

We'll start using `R` as a glorified calculator to get familiar with the syntax and workflow.

:::{.fragment}
```{r}
# Addition
5 + 2
```
:::

:::{.fragment}
```{r}
# Subtraction
5 - 2
```
:::

:::{.fragment}
```{r}
# Multiplication
5 * 2
```
:::

:::{.fragment}
```{r}
# Division
5 / 2
```
:::

## Basic operations (Part 2)

We'll start using `R` as a glorified calculator to get familiar with the syntax and workflow.

:::{.fragment}
```{r}
# Power
2 ^ 2
2 ** 2
```
:::

:::{.fragment}
```{r}
# Square Root
sqrt(16)
```
:::

:::{.fragment}
```{r}
# Absolute Value
abs(-10)
```
:::

## Function arguments

Functions are code commands to perform specific tasks. Functions are `R` objects containing multiple interrelated statements that are run together.

`Base R` and packages provide a huge variety of predetermined functions, but the user can create custom functions as well.

- The values we pass into a function are called *arguments*.
- Some functions only need one argument, usually an `R` object or a scalar value such as `sqrt(16)`.
- Others need several arguments, and their order can matter if we do not name them.
- Writing argument names explicitly usually makes the code easier to read and safer to edit.

## Positional and named arguments

If we do not name arguments, `R` reads them by position.

- Positional arguments depend on order.
- Named arguments make the meaning of each input explicit.
- Named arguments are especially useful when a function has several options.

:::{.fragment}
```{r}
round(3.14159, 2)
round(x = 3.14159, digits = 2)
seq(1, 10, 2)
seq(from = 1, to = 10, by = 2)
```
:::

## Objects and overwriting

A core habit in `R` is assigning results and values to objects.

- The assignment operator is `<-`.
- Alternatively `=` can also be used, but `<-` is more common in `R`.
- An object stores a result under a name.
- Overwriting changes what that name refers to.

:::{.fragment}
```{r}
x <- 5
x

x <- 5 + 2
x
```
:::

- `R` is case sensitive, so `x` and `X` are different objects.
- As corollary, it is advisable to name `R` objects with lowercase letters to avoid confusion.

## Showing results with `print()`

Most of the time, typing an object name and running the code is enough to display its contents.

- `print()` explicitly tells `R` to display an object or result.
- It is useful when you want to make the output step explicit.
- It becomes especially useful inside functions, loops, and more complex scripts.

:::{.fragment}
```{r}
x <- 5 + 2
x
print(x)
print("Hello, class")
```
:::

## Vectorized operations (Part 1)

A **vector** is `R`'s most basic data structure, capable of storing multiple entries of the same type.

The `c()` function combines elements into a vector.

Many `R` functions then work on the whole vector directly, which is one of the language's key habits.

:::{.fragment}
```{r}
vector_num <- c(5, 2, 10)
vector_num
abs(vector_num)
vector_num + 5
```
:::

- Single element vectors are called *atomic* vectors, and they are the building blocks of more complex data structures.

## Vectorized operations (Part 2)

Some functions return a single value when applied to a vector, while others return a vector of the same length.

:::{.fragment}
```{r}
# Minimum, maximum, mean, and square root of a vector
min(vector_num)
max(vector_num)
mean(vector_num)
sqrt(vector_num)
```
:::

## Text objects and text vectors

Character strings are text values, and they can also be stored in vectors.

:::{.fragment}
```{r}
var_text <- "UCAB"
var_text
vector_text <- c("Course", "data", "science")
vector_text
var_text2 <- "2025"
paste(var_text, var_text2)
paste(vector_text, collapse = " ")
combined <- paste(c(vector_text, var_text, var_text2), collapse = " ")
combined
```
:::

## Packages and built-in data

Packages are `R` extensions that add functionalities, allowing users to add methods and tools very quickly.

`library()` loads an installed package into the session, making its functions and datasets available.

Some packages come with other packages as dependencies, which are automatically installed but need to be loaded separately.

:::{.fragment}
```{r eval=FALSE}
# Install package
install.packages("dslabs")

# Load packages
library(dslabs)
```
:::

Some packages come with built-in datasets that can be loaded with `data()`. The `USArrests` dataset is one of the base datasets that come with `R`.

- The `?` operator gives access to the help system, displaying documentation for functions and datasets.

:::{.fragment}
```{r eval=FALSE}
data()
base <- USArrests
?USArrests
View(base)
```
:::

The practical rule is simple: install once, load when needed, and use the help system constantly.

## Installing many packages at once

The same idea also works at larger scale.

:::{.fragment}
```{r eval=FALSE}
packages <- c(
  "dslabs", "tidyverse", "dplyr", "ggplot2", "swirl", "readxl", "openxlsx",
  "htmltools", "readstata13", "haven", "stringr", "purrr",
  "data.table", "foreign", "lubridate", "ggrepel", "rvest",
  "pdftools", "ggthemes", "gridExtra", "scales",
  "RColorBrewer", "usmap", "maps", "egg", "gtools", "VennDiagram"
)
install.packages(packages)
```
:::

More complex scripts can be built to check if a package is already installed before trying to install it, but the above code is usually enough for most cases.

## Intervention Space {.intervention-slide}

Question to ponder.

:::{.fragment}
- Could installing and loading multiple packages at once cause any issues?
:::

:::{.fragment}
- *Answer:* Yes, it could lead to conflicts between packages if they have functions with the same names. It can also make it harder to troubleshoot issues if something goes wrong during installation or loading. Additionally, it may take more time to install and load a large number of packages at once, especially if some of them have many dependencies.
:::


## Data types begin with `class()`

Different objects in `R` have different classes.

The `class()` function tells us what kind of object `R` thinks it is reading.

:::{.fragment}
```{r}
a <- 2
class(a)
a <- "2"
class(a)
```
:::

## The `murders` dataset (Part 1)

From here, the lecture uses the `murders` dataset from `dslabs` as the first real case.

The `data()` function loads a dataset into the session. This dataset shows that data frames are the standard way to store datasets in `R`.

- A data frame is a table.
- Rows represent observations.
- Columns represent variables.
- Different columns can have different data types.

## The `murders` dataset (Part 2)

The `str()` function gives a compact structural summary, and `head()` shows the first rows.

:::{.fragment}
```{r}
# Load library and packages
library(dslabs)
data(murders)

# Inspect the dataset
class(murders)
str(murders)
head(murders)
```
:::

## Accessing columns with `$`

The `$` accessor lets us extract one variable from a data frame.

- This is a natural move from the full table to one column.
- The result is often a vector.
- Once extracted, we can inspect or reuse it like any other object.

:::{.fragment}
```{r}
murders$population
pop <- murders$population
class(pop)
length(pop)
```
:::

## `names()`, `colnames()`, and `length()`

These are basic inspection functions.

- `names()` and `colnames()` return variable names.
- `length()` returns the number of entries in a vector.

:::{.fragment}
```{r}
names(murders)
colnames(murders)
length(pop)
```
:::

## Coercion and logical values

- Coercion is `R` trying to force conversion between data types.
- That flexibility can also create silent mistakes.
- Logical values are a separate data type built from `TRUE` and `FALSE`.

:::{.fragment}
```{r}
vector <- c(1, 2, "car")
vector
class(vector)
3 == 2
z <- 3 == 2
z
```
:::

## Factors and lists (Part 1)

Not every object behaves like a simple vector.

Factors are `R`'s way to store categorical data. Preferably used when values belong to a fixed set of categories. They are stored as integers with labels, which can be useful for modeling but also requires care when manipulating.

The `levels()` function reveals the categories stored inside a factor.

:::{.fragment}
```{r}
class(murders$region)
levels(murders$region)
```
:::

## Factors and lists (Part 2)

`list()` creates an object that can store components of different types.

List elements don't need to be of the same size or type, and they can be accessed with `$` or `[[ ]]`.

:::{.fragment}
```{r}
record <- list(
  name = "John Doe",
  student_id = 1234,
  grades = c(95, 82, 91, 97, 93),
  final_grade = "A"
)
class(record)
record$student_id
```
:::

## Using `sort()` and `order()` (Part 1)

Which state has the highest number of firearm murders in 2010, and which has the lowest?

The `sort()` function returns ordered values. The `order()` function returns the positions that would sort those values.

- By default, both functions sort from smallest to largest.
- `decreasing = TRUE` changes the order from largest to smallest.

:::{.fragment}
```{r}
sort(murders$total)
x <- c(31, 4, 15, 92, 65)
sort(x)
sort(x, decreasing = TRUE)
```
:::

## Using `sort()` and `order()` (Part 2)

The `sort()` function returns ordered values. The `order()` function returns the positions that would sort those values.

:::{.fragment}
```{r}
index <- order(x)
index
order(x, decreasing = TRUE)
x[index]
```
:::

## Applying `order()` to the case study

:::{.fragment}
```{r}
index <- order(murders$total)
murders$state[index]
```
:::

## `which.max()` and `which.min()`

When only the extreme case matters, a more direct route is useful.

These functions return the position of the largest or smallest value in a vector.

:::{.fragment}
```{r}
max(murders$total)
min(murders$total)
which.max(murders$total)
which.min(murders$total)
index <- which.max(murders$total)
murders$state[index]
index <- which.min(murders$total)
murders$state[index]
```
:::

## Named vectors

We can also add names to vectors to make them easier to read and subset.

The `names()` function can also assign names to an existing vector, which makes indexing more readable.

Naming vectors does not change their class, but it adds a layer of information that can be useful for subsetting and interpretation.

:::{.fragment}
```{r}
codes <- c(380, 124, 818)
codes
country <- c("italy", "canada", "egypt")
codes <- c("italy" = 380, "canada" = 124, "egypt" = 818)
codes
class(codes)
names(codes)
```
:::

## `seq()` and `:`

These are two common ways to build sequences.

- `seq()` is flexible because it lets us set the step size.
- `:` is compact and useful for consecutive integers.

:::{.fragment}
```{r}
seq(1, 10)
seq(1, 10, by = 2)
1:10
class(1:10)
class(seq(1, 10, 0.5))
```
:::

## Subsetting vectors

Subsetting tells `R` which elements we want to keep or extract.

- `[]` returns a subset while preserving the original structure.
- `[[]]` extracts a single element from a list or data frame.
- With vectors, `[]` usually returns another vector.
- With lists, the difference matters because the returned object is not the same.

:::{.fragment}
```{r}
codes[2]
class(codes[2])
codes[c(1, 3)]
codes["italy"]
my_seq <- 1:100
my_seq_even <- seq(2, 100, by = 2)
```
:::

## Subsetting lists with `[]` and `[[]]`

Lists are a good place to see the difference clearly.

- `record["student_id"]` keeps the result as a list.
- `record[["student_id"]]` extracts the value itself.
- Looking at `class()` makes the structural difference visible.

:::{.fragment}
```{r}
record["student_id"]
```
:::

:::{.fragment}
```{r}
class(record["student_id"])
```
:::

:::{.fragment}
```{r}
record[["student_id"]]
```
:::

:::{.fragment}
```{r}
class(record[["student_id"]])
```
:::

## Coercion and `NA` (Part 1)

We can try to make `R` interpret values as a different type, and that is called coercion.

The `as.character()` and `as.numeric()` functions explicitly convert one type into another.

- Sometimes conversion works cleanly.
- Sometimes it produces `NA`.
- That is usually a sign that a value could not be interpreted as the requested type.

:::{.fragment}
```{r}
x <- c(1, "canada", 3)
x
class(x)
x <- 1:5
y <- as.character(x)
y
```
:::

## Coercion and `NA` (Part 2)

:::{.fragment}
```{r}
y <- as.numeric(y)
y
country_num <- as.numeric(country)
country_num
x <- c("1", "b", "3")
as.numeric(x)
```
:::

## Logical indexing

Logical indexing lets us answer more interesting questions.

This is the first time the lecture uses a logical vector as a filter.

- We build a condition.
- `R` evaluates it for every entry.
- The resulting logical vector can be used to subset another vector.

:::{.fragment}
```{r}
data(murders)
murder_rate <- murders$total / murders$population * 100000
ind <- murder_rate <= 0.71
ind
murders$state[ind]
```
:::

## Combining conditions with `&`

The operator `&` means both conditions must be true at the same time.

:::{.fragment}
```{r}
west <- murders$region == "West"
safe <- murder_rate <= 1
ind <- safe & west
murders$state[ind]
```
:::

## `which()`, `match()`, and `%in%` (Part 1)

These three tools all help us locate values, but they do not return the same type of result.

- `which()` returns positions where a condition is true.
- `match()` returns the position of requested values inside another vector.
- `%in%` returns logical values.

:::{.fragment}
```{r}
ind <- which(murders$state == "California")
ind
murder_rate[ind]
murder_rate[murders$state == "California"]
ind <- match(c("New York", "Florida", "Texas"), murders$state)
ind
```
:::

:::{.fragment}
```{r}
murder_rate[ind]
a <- 1:10
b <- c(3, 15, 7)
b %in% a
```
:::

## `which()`, `match()`, and `%in%` (Part 2)

```{r}
murder_rate[ind]
a <- 1:10
b <- c(3, 15, 7)
b %in% a
```

## Intervention Space {.intervention-slide .incremental}

Answer these before the lecture ends.

- Why does `murders$population` become a vector when extracted from the data frame?
- Why is `which.max()` more useful than `sort()` for some questions but not for all of them?
- Why does `as.numeric(country)` return `NA` values?
- What is the conceptual difference between `which()`, `match()`, and `%in%`?

## Main Message {.main-message}

- *Data Science is a multidisciplinary field that combines analytical thinking, programming, and storytelling to extract insights from data.*
- *This course serves as a practical introduction to the field through the use of `R` and `RStudio` as main tools to think, code, and inspect data.*
