---
title: "Lecture 03"
subtitle: "Combining, Importing, and Cleaning Data"
format: revealjs
---

## Main Message {.main-message}

*Real data work rarely starts with one clean table.*

*Before analysis can be trusted, we often need to import files, combine related tables through keys, and clean text values into usable variables.*

# From One Table To Several Tables

## Why this comes after `dplyr`

In Lecture 02, we worked mainly with one tidy data frame at a time.

- `select()` chose columns.
- `filter()` chose rows.
- `mutate()` created variables.
- `summarise()` and `group_by()` produced grouped summaries.
- `pivot_longer()` and `pivot_wider()` reshaped a table.

Those tools are most useful once the data is already in a workable table.

Many projects start one step earlier: the information is split across files, tables, and text fields that need preparation.

## Three preparation tasks

This lecture adds three common preparation tasks:

1. Combining tables that share a key.
2. Importing data from external files.
3. Cleaning imported text values before analysis.

The goal is not to memorize every possible import or cleaning function. The goal is to recognize which operation the data needs.

```{r}
#| include: false
library(tidyverse)
library(dslabs)

data(murders)
```

## A key is how tables recognize each other

A *key* is a variable that identifies matching observations across tables.

In this example, both tables describe U.S. states. The `state` column is the key we can use to connect them.

::: fragment
```{r}
# Inspect the analysis table
murders %>%
  select(state, region, population, total) %>%
  head()
```
:::

::: fragment
```{r}
# Inspect the lookup table
results_us_election_2016 %>%
  select(state, electoral_votes) %>%
  head()
```
:::

## Row position is not a key

Two tables can contain the same kind of observation without storing rows in the same order.

::: fragment
```{r}
# Check whether state names appear in the same row order
identical(murders$state, results_us_election_2016$state)
```
:::

`FALSE` means we should not paste these columns together by position. We need a join that matches rows using `state`.

## Intervention Space {.intervention-slide}

Question to ponder.

::: fragment
- If two tables both have a `state` column, what should we check before joining?
:::

::: fragment
- *Answer:* Check that the columns have the same meaning, compatible values, and the expected level of uniqueness. Same name alone is not enough.
:::

# Joins

## `left_join()` adds columns by matching keys

`left_join()` keeps all rows from the first table and adds matching columns from the second table.

::: fragment
```{r}
# Add electoral votes to the murders table
murders_election <- murders %>%
  left_join(results_us_election_2016, by = "state")
```
:::

::: fragment
```{r}
# Keep only a few columns for inspection
murders_election %>%
  select(state, population, total, electoral_votes) %>%
  head()
```
:::

## A small join example

The differences among joins are easier to see with smaller tables.

::: fragment
```{r}
# Main table
tab1 <- murders %>%
  slice(1:6) %>%
  select(state, population)

tab1
```
:::

::: fragment
```{r}
# Lookup table with only some matching states
tab2 <- results_us_election_2016 %>%
  slice(c(1:3, 5, 14, 44)) %>%
  select(state, electoral_votes)

tab2
```
:::

## Four mutating joins

Mutating joins add columns. They differ in which rows they keep.

::: fragment
```{r}
# Keep every row from tab1
left_join(tab1, tab2, by = "state")
```
:::

::: fragment
```{r}
# Keep every row from tab2
right_join(tab1, tab2, by = "state")
```
:::

## Four mutating joins

The other two common mutating joins keep only matches or keep everything.

::: fragment
```{r}
# Keep only states that appear in both tables
inner_join(tab1, tab2, by = "state")
```
:::

::: fragment
```{r}
# Keep states from either table
full_join(tab1, tab2, by = "state")
```
:::

## Filtering joins

Filtering joins use another table to decide which rows to keep. They do not add columns.

::: fragment
```{r}
# Rows in tab1 that have a match in tab2
semi_join(tab1, tab2, by = "state")
```
:::

::: fragment
```{r}
# Rows in tab1 that do not have a match in tab2
anti_join(tab1, tab2, by = "state")
```
:::

## Intervention Space {.intervention-slide}

Question to ponder.

::: fragment
- Which join would you use if you want every row from your main dataset, plus whatever matches from a lookup table?
:::

::: fragment
- *Answer:* Use `left_join()` with the main dataset on the left.
:::

# Binding And Set Operations

## Binds do not match by key

Binding combines objects by position or compatible structure. It does not search for matching key values.

::: fragment
```{r}
# Bind vectors as columns
bind_cols(a = 1:3, b = 4:6)
```
:::

::: fragment
```{r}
# Bind data frames as rows
bind_rows(
  tibble(state = c("Alabama", "Alaska"), group = "first"),
  tibble(state = c("Arizona", "Arkansas"), group = "second")
)
```
:::

## Joins versus binds

Use a join when the question is:

- Which rows in table A match rows in table B?
- Which columns should be added based on a key?
- Which rows have no match?

Use a bind when the question is:

- Can these rows be stacked because they have compatible columns?
- Can these columns be placed side by side because the row order already means something?

When in doubt, look for the key first.

## Set operations compare membership

Set operations ask what values or rows are shared, combined, or different.

::: fragment
```{r}
# Values in both vectors
intersect(1:10, 6:15)
```
:::

::: fragment
```{r}
# Values in the first vector but not the second
setdiff(1:10, 6:15)
```
:::

::: fragment
```{r}
# Same values, different order
setequal(1:5, 5:1)
identical(1:5, 5:1)
```
:::

# Importing Files

## Scripts should remember how data arrived

Using built-in datasets is useful for learning, but most projects start with files.

A reproducible import records:

- where the file came from;
- which function read it;
- which options were used;
- which object stores the result.

Point-and-click importing can help explore a file, but the final workflow should live in a script.

## Reading a CSV

The `dslabs` package includes a CSV version of the `murders` data. We can locate it without copying it into the project folder.

::: fragment
```{r}
# Build a reproducible path to a package file
csv_path <- system.file("extdata", "murders.csv", package = "dslabs")

csv_path
```
:::

::: fragment
```{r}
# Read the CSV into R
imported_murders <- read_csv(csv_path, show_col_types = FALSE)

imported_murders %>%
  select(state, population, total) %>%
  head()
```
:::

## Inspect imported types

After import, always check whether R interpreted the variables as expected.

::: fragment
```{r}
# Compare a few imported column classes
imported_murders %>%
  select(state, population, total) %>%
  summarise(across(everything(), ~ class(.x)[1]))
```
:::

This habit matters because imported numbers, dates, and categories often arrive as text.

## Intervention Space {.intervention-slide}

Question to ponder.

::: fragment
- Why is a written import command more reproducible than clicking the RStudio Import button?
:::

::: fragment
- *Answer:* The command preserves the path, function, arguments, and object name. Someone else can rerun it later and see exactly how the data entered R.
:::

# Cleaning Imported Text

## Imported values can look numeric but behave like text

Some values arrive with commas, labels, percent signs, or other characters.

::: fragment
```{r}
# A small example that mimics imported text
messy_counts <- tibble(
  state = c("California", "Texas", "Florida"),
  population_text = c("37,253,956 people", "25,145,561 people", "18,801,310 people"),
  murders_text = c("1,257", "805", "987")
)

messy_counts
```
:::

::: fragment
```{r}
# The population column is text, not numeric
class(messy_counts$population_text)
```
:::

## Detect the pattern

`str_detect()` checks whether a character vector contains a pattern.

::: fragment
```{r}
# Which population values contain commas?
str_detect(messy_counts$population_text, ",")
```
:::

::: fragment
```{r}
# Do any values contain commas?
any(str_detect(messy_counts$population_text, ","))
```
:::

## Replace or parse

`str_replace_all()` changes matching text. `parse_number()` extracts numeric content from text.

::: fragment
```{r}
# Remove commas but keep the rest of the text
messy_counts %>%
  mutate(population_no_commas = str_replace_all(population_text, ",", ""))
```
:::

::: fragment
```{r}
# Extract usable numbers
clean_counts <- messy_counts %>%
  mutate(population = parse_number(population_text),
         total = parse_number(murders_text))

clean_counts
```
:::

## Confirm the cleaned type

Cleaning is not complete until we check the result.

::: fragment
```{r}
# The cleaned population is numeric
class(clean_counts$population)
```
:::

::: fragment
```{r}
# Now we can compute a rate
clean_counts %>%
  mutate(rate = total / population * 100000) %>%
  select(state, rate)
```
:::

## Cleaning can also fix keys

Text cleaning is useful before joins when keys are written in different cases.

::: fragment
```{r}
# A lookup table with lower-case state names
region_lookup <- tibble(
  state = c("california", "texas", "florida"),
  short_region = c("West", "South", "South")
)
```
:::

::: fragment
```{r}
# Standardize the key before joining
region_lookup %>%
  mutate(state = str_to_title(state)) %>%
  left_join(clean_counts, by = "state")
```
:::

## Intervention Space {.intervention-slide}

Question to ponder.

::: fragment
- Why can `as.numeric("1,234")` produce `NA`, while `parse_number("1,234")` returns `1234`?
:::

::: fragment
- *Answer:* `as.numeric()` expects a clean numeric string. `parse_number()` is designed for imported text and extracts the numeric part before converting it.
:::

# Practice

## Join, import, clean

Use the tools together:

1. Import a CSV into R.
2. Inspect the column types.
3. Clean any text values that should become numbers.
4. Join a lookup table when extra columns are needed.
5. Compute a summary only after the structure and types make sense.

::: fragment
```{r}
# A compact example using the cleaned toy table
clean_counts %>%
  mutate(rate = total / population * 100000) %>%
  left_join(region_lookup %>% mutate(state = str_to_title(state)), by = "state") %>%
  group_by(short_region) %>%
  summarise(avg_rate = mean(rate), .groups = "drop")
```
:::

## Main Message {.main-message}

*Real data work rarely starts with one clean table.*

*The practical workflow is: import the data, combine related tables through keys, clean values that arrived as text, and only then summarize or analyze.*
